---
title: "Annotating and Filtering"
author: "Mark Dunning"
date: "24 June 2016"
output: html_document
---

# Outline

We've 

## Introduction to annovar

***Important:- annovar license***

`annovar` is free for academice use. However, you need to supply a valid academic email address in order to download the software. If you wish to use `annovar` after the course. You will have to fill out the form on the [annovar](http://annovar.openbioinformatics.org/en/latest/user-guide/download/) website. 


`annovar` provides "gene-based", region-based" and "filter-based" annotation. 

- *Gene-based*: whether variants cause a protein-coding change. If so, it will report the affected amino acids
    + "RefSeq genes, UCSC genes, ENSEMBL genes, GENCODE genes, AceView genes, or many other gene definition systems."
- *Region-based*: look for overlaps with pre-defined genomic regions
    + e.g.  "conserved regions among 44 species, predicted transcription factor binding sites, segmental duplication regions, GWAS hits, database of genomic variants, DNAse I hypersensitivity sites, ENCODE H3K4Me1/H3K4Me3/H3K27Ac/CTCF sites, ChIP-Seq peaks, RNA-Seq peaks, or many other annotations on genomic intervals"
- *Filter-based*: looks for variants in pre-defined databases. Not only the genomic location is taken into account, but also the change from reference to alternate allele
    + dbSNP, 1000 Genome Project, NHLBI-ESP 6500 exomes or Exome Aggregation Consortium, 
  
annovar relies on pre-downloaded files so all queries are done offline. For the course, we have downloaded some of these files to the directory `/home/participant/Course_Materials/ref_data/annovar/humandb`. Each annotation file has a genome version associated with it. For our purposes we are going to be using `hg19`. 

```{}
ls -l ../ref_data/annovar/humandb
```


## Preparing files for annovar

Some of the commands in `annovar` require a specific input format. Fortunately, they have provided a script that can convert from popular file formats to their format. 


First we need to convert our `.vcf` file using a script from annovar 'convert2annovar.pl`. 

```{}
/home/dunnin01/software/annovar/convert2annovar.pl -format vcf4old HCC1143_vs_HCC1143_BL.flagged.muts.vcf > HCC1143_vs_HCC143_BL.flagged.muts.avinput
```

```{r}
cat(system("head HCC1143_vs_HCC1143_BL.flagged.muts.avinput",intern=TRUE),sep="\n")
```

"The first five space or tab delimited fields are Chromosome ("chr" prefix is optional), Start, End, Reference Allelel, Alternative Allele. The rest of the columns are completely optional."

    
### Region-based Annotation

Gene-based annotation is the default operation in annovar. The following command will annotate the file we have just prepared against the reference file located in `../ref_data/annovar/humandb/hg19_refGene.txt`.


```{}
/home/dunnin01/software/annovar/annotate_variation.pl -buildver hg19 HCC1143_vs_HCC1143_BL.flagged.muts.avinput /home/dunnin01/software/annovar/humandb
```

Two files are created:- `HCC1143_vs_HCC1143_BL.flagged.muts.vcf.variant_function` and `HCC1143_vs_HCC1143_BL.flagged.muts.vcf.variant_function`

```{}
cat(system("head HCC1143_vs_HCC1143_BL.flagged.muts.avinput.exonic_variant_function",intern=TRUE),sep="\n")
```

```{}
cat(system("head HCC1143_vs_HCC1143_BL.flagged.muts.avinput.variant_function",intern=TRUE),sep="\n")
```

******

Exercise

- Read the variant function results into R
- How many of each type of variant do we have?
- Read the exonic variant function file into R
    + how many nonsyonymous SNVs do we have?
    + how many synonumous? etc
    
******

```{r}
varfun <- read.delim("HCC1143_vs_HCC1143_BL.flagged.muts.avinput.variant_function",header=FALSE)

```


```{r}
exonfun <- read.delim("HCC1143_vs_HCC1143_BL.flagged.muts.avinput.exonic_variant_function",header=FALSE)
```

### Region-based annotation

The simplest form of region annotation is probably when we want to look at which "cytoband" each variant belongs to. This uses a reference file that contains the coordinates for each cytoband in a given genome version. 



```{}
head ../ref_data/annovar/humandb/hg19_cytoBand.txt
```

```{r echo=FALSE,comment=NA}
cat(system("head ../ref_data/annovar/humandb/hg19_cytoBand.txt",intern=TRUE),sep="\n")
```

If didn't already have such a file it could be downloaded using:- 

```{}
/home/dunnin01/software/annovar/annotate_variation.pl -buildver hg19 -downdb cytoBand ../ref_data/annovar/humandb/
```

To perform the annotation, we need to specify `-regionanno`, to perform region-annotation and set the `dbtype` to `cytoBand`. 

```{}
/home/dunnin01/software/annovar/annotate_variation.pl -regionanno -build hg19 -dbtype cytoBand  HCC1143_vs_HCC1143_BL.flagged.muts.avinput ../ref_data/annovar/humandb/
``` 

This should result in a file being created with `.cytoBand` added as a file extension to your input file (e.g. `HCC1143_vs_HCC1143_BL.flagged.muts.avinput`). An extra column has been added to indicate which chromosome band each variant is located on. In the event of a variant spanning multiple bands (obviously not in this case), the multiple bands would be listed here.

```{}
head HCC1143_vs_HCC1143_BL.flagged.muts.avinput.cytoBand
```

```{r}
cat(system("head HCC1143_vs_HCC1143_BL.flagged.muts.avinput.hg19_cytoBand",intern=TRUE),sep="\n")
```

In another example, we could annotation against locations that have been reported in previous GWAS studies 

```{}
/home/dunnin01/software/annovar/annotate_variation.pl -regionanno -build hg19 -dbtype gwasCatalog HCC1143_vs_HCC1143_BL.flagged.muts.avinput ../ref_data/annovar/humandb
```

```{}
cat(system("head HCC1143_vs_HCC1143_BL.flagged.muts.avinput.hg19_gwasCatalog",intern=TRUE),sep="\n")
```


### Filter-based annotation

Unlike the annotation performed in the previous section, which was based purely on the genomic location, *filter-based* annotation also takes the base-change into account. 

Several types of database are available

- Frequency of variants in whole-genome data

- Frequency of variants in whole-exome data

- Functional prediction of variants in whole-genome data

- Functional prediction of variants in whole-exome data

- Disease-specific variants


For example, if we want to see which of our variants are present a in database of known (or common) SNPs we could use the `dbSNP` database. 

Now we select the `filter` option in annovar with a `dbtype` of `snp129`
```{}
/home/dunnin01/software/annovar/annotate_variation.pl -filter -buildver hg19 -dbtype snp129  HCC1143_vs_HCC1143_BL.flagged.muts.avinput ../ref_data/annovar/humandb/
```

This time we get two files created; `HCC1143_vs_HCC1143_BL.flagged.muts.avinput.hg19_snp129_dropped` and `HCC1143_vs_HCC1143_BL.flagged.muts.avinput.hg19_snp129_filtered`. The difference being that the `.._dropped` file contains details of all your variants that ***are*** in the database

```{}
head HCC1143_vs_HCC1143_BL.flagged.muts.avinput.hg19_snp129_dropped
```

```{r}
cat(system("head HCC1143_vs_HCC1143_BL.flagged.muts.avinput.hg19_snp129_dropped",intern=TRUE),sep="\n")
```

Thus, the `.._filtered` file contains all the variants ***not found*** in the database. 

We can perform annotation against 1000 genomes variants with the following:-

```{}
../software/annovar/annotate_variation.pl -filter -dbtype 1000g2014oct_eur -buildver hg19 HCC1143_vs_HCC1143_BL.flagged.muts.avinput ../ref_data/annovar/humandb/
```

As before, `..._dropped` and `...filtered` files are produce. Again the `.._dropped` file contains variants that were found in the database. However, rather than printing and ID for the variant, it prints the minor allele frequency that is observed for that variant.

```{}
head HCC1143_vs_HCC1143_BL.flagged.muts.avinput.hg19_EUR.sites.2014_10_dropped
```

```{r echo=FALSE}
cat(system("head HCC1143_vs_HCC1143_BL.flagged.muts.avinput.hg19_EUR.sites.2014_10_dropped",intern=TRUE),sep="\n")
```

We can also provide a minor allele frequency (maf) threshold. This will mean that only variants with a maf greater than the threshold will be printed to the `..._dropped` file.

```{}
../software/annovar/annotate_variation.pl -filter -dbtype 1000g2014oct_eur -buildver hg19 HCC1143_vs_HCC1143_BL.flagged.muts.avinput ../ref_data/annovar/humandb/ -maf 0.05
```

The commands above annotated against variants derived from european indivduals. If we have individuals of a different ethnicity, we probably want to use a different cohort. In this case, we annotate against frequencies observed in asian individuals.

```{}
../software/annovar/annotate_variation.pl -filter -dbtype 1000g2014oct_afr -buildver hg19 HCC1143_vs_HCC1143_BL.flagged.muts.avinput ../ref_data/annovar/humandb/
```

Clinical variants.....

```{}
../software/annovar/annotate_variation.pl -filter -dbtype clinvar_20160302 -buildver hg19 HCC1143_vs_HCC1143_BL.flagged.muts.avinput ../ref_data/annovar/humandb/
```


## Combining multiple annotations and filters

Up to this point, we have amassed quite a collection of tables of our set of variants annotated / filtered against various databases. A particularly appealing feature of `annovar` is that we can define protocols to combine multiple filters and collect the results in the same table. 

The `-protocol` argument is used to list what types of annotation / filtering you want to be applied. With the `-operation` argument we have to specify whether each annotation is a gene (`g`), region (`r`) or filter (`f`) -based annotation. You many need to take care to provide the same number of names in both the `-protocol` and `-operation` arguments.

```{}
../software/annovar/table_annovar.pl HCC1143_vs_HCC1143_BL.flagged.muts.avinput ../ref_data/annovar/humandb/ -buildver hg19 -out myanno -remove -protocol refGene,cytoBand,gwasCatalog,snp129,1000g2014oct_eur,1000g2014oct_afr,clinvar_20160302 -operation g,r,r,f,f,f,f -nastring . -csvout
```

The file `myanno.hg19_multianno.csv` is created. By specifying the `-remove` argument in the command, any tables that are created as part of the process are removed. This `.csv` file can now be viewed in Excel, or LibreOffice in our case (right-click on the )

```{}
head myanno.hg19_multianno.csv
```

```{r, echo=FALSE,comment=NA}
cat(system("head myanno.hg19_multianno.csv",intern=TRUE),sep="\n")
```




## Annotating using snpEFF

